Reasoning About a Machine with Local Capabilities

### 2018-05-01 toplas/paper.tex
Wrote the rest of the malloc section text

Interestingly, this idea is a variant of a notion of public/private future
worlds that has been previously used in the literature (see
\sectionname~\ref{sec:discussion}). However, temporary regions are new in our setting
and there is an interesting interplay with the recursiveness of the worlds: for
a temporary region, the predicate $H \in \Worlds \rightarrow
\powerset{\MemSegments}$ (which defines the safe memory segments in the
current world) is only required to be monotone w.r.t.\ public
future worlds (i.e.\ safe memory segments remain safe in public future worlds),
but for permanent regions, it must be monotone w.r.t.\ private future worlds.
As a consequence, the memory for a permanent region may not contain local
capabilities (as their safety would be broken in private future worlds), which
in turn implies that only local capabilities may have write-local permission (a
general sanity requirement when using local capabilities\footnote{In fact, local
  capabilities become useless as soon as the adversary has access to a single
  global, write-local capability.})\footnote{For explanation purposes, this
  discussion ignores certain ways to allow for local capabilities in a permanent
  region, for example, by not requiring that they are valid or requiring that
  they are local versions of valid global capabilities.}.
\end{toplas}
\subsection{Worlds}
\begin{figure}[htb]
  \centering
  \includegraphics{w11}

@@

\begin{toplas}
  \section{Malloc and macros}
  \label{sec:malloc-and-macros}
With the semantic model defined, we would like to show its usefulness by proving
the correctness of a series of examples that use the calling convention and rely
on well-bracketedness and local-state encapsulation. However, the programs that
run on our capability are assembly programs and even program examples that would
be small in a high-level language, they become big and unintelligible at this
low level. In order to make our program examples intelligible, we introduce a
series of low-level abstractions in the form of macros.

In order to make the program examples realistic, we allow them to allocate fresh
memory by invoking a trusted malloc. Our results are not specific to
any one implementation of malloc, so rather than providing one implementation,
we provide a specification that malloc must satisfy.

\subsection{Malloc specification}
  \itoplassug{Suggestion: add the malloc specification}
Before we give the malloc specification, we provide some intuition about what
the malloc specification should look like.

Malloc is assumed to be part of the trusted computing base, as mentioned in
\sectionname~\ref{sec:capab-mach-with}. A program with access to malloc should
be able to retain this access even if it invokes code it does not trust. In
order to achieve this, a capability for malloc must be global. It is also
important that malloc remains trustworthy to all entities on the machine
throughout execution. That is, malloc should encapsulate its local state and
not depend on any external code. To make sure that malloc cannot be tampered
with through the capability for invoking malloc, we also require the capability
to have enter permission.

The requirements so far prevent direct tampering in the sense that the
capability intended for invoking malloc only can be used for invocation. This
does, however, not prevent the implementation of malloc from colluding with
other code. Specifically, malloc could depend on the state of a piece of memory
not just under the control of malloc. We also need to make sure that malloc
actually behave as intended. That is when malloc is invoked with a length
argument in the correct register, then it should actually return a capability
for a fresh piece of memory of that size. It should be fresh in the sense that
malloc hasn't given out a capability for any part of that memory before and
won't do it in the future. Finally, we also need to make sure that malloc
behaves correctly when given a nonsensical length argument such as a negative
integer or a capability. These requirements relate to the behaviour of malloc,
so we cannot hope to specify them syntactically. Instead we will use the
semantic model defined in  Section~\ref{sec:logical-relation} to express this
part of the specification.

In order to express the semantic requirements on malloc, the specification will
require the existence of some initial region for malloc. Malloc does not
disappear from the system, so the malloc region must permanent. This also means
that the malloc specification should guarantee that use of malloc is safe at any
point of execution and no matter who has invoked it. To this end, the
requirement on the initial malloc region should also hold true in any private
future world.

The implementation of malloc should not depend on any memory outside the
implementation of malloc. This means that malloc should not have any
capabilities in its internals that govern other programs. The malloc region's
memory interpretation function determines what malloc may contain, so we require
it to allow the same memory segments whether it is with respect to some world
with the malloc region in it or a world that just contains the malloc region.
Malloc should retain this property throughout execution, so it must hold true
for any private future region of the initial malloc region.

The malloc specification also dictates what malloc should do when invoked. We
specify the two cases where malloc is invoked with a valid and an invalid length
argument, respectively. When malloc is invoked with an invalid length argument,
that is a negative integer or a capability, we require malloc to simply fail.
This part of the specification does not rely on the malloc region as we do not
need to make sure that the memory invariants are respected. 
On the other hand,
\begin{definition}[Malloc Specification]
  \label{spec:malloc}
  $c_\malloc$ satisfies the specification for malloc iff
  \[  
    \begin{array}{l}
      c_\malloc = ((\entry,\glob),\_,\_,\_) \land\\
      \exists \iota_{\malloc,0} \ldotp \\
      \quad \iota_{\malloc,0}.v = \perma \; \land \\
      \quad \left(
        \begin{array}{l}
          \forall \iota' \futurestr \iota_{\malloc,0}, W,i \ldotp\\
          \quad W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
        \end{array}
        \right) \wedge \\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
        \quad \forall i, n, \size \in \nats \ldotp \forall
        w_{\var{ret}} \in \Words \ldotp \\
        \quad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
        \quad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
        \qquad \memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
        \qquad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
        \qquad \Rightarrow \\
        \qquad \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
        \qquad \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
        \qquad \qquad \quad \Phi \step[j] \Phi' \land \\
        \qquad \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
        \qquad \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
        \qquad \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
        \qquad \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
        \qquad \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
        \qquad \qquad \quad \size - 1 = e'-b'
      \end{array}
      \right) \land\\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp\\
        \quad(\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
        \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
        \qquad \Rightarrow \exists j \in \nats \ldotp \Phi \step[j] \failed
      \end{array}
        \right)
    \end{array}
  \]
\end{definition}

\itoplassug{Write something about initial capabilities to malloc. We will always
assume that capabilities for malloc satisfy the malloc specification. The malloc
specification prevents malloc from giving away internal capabilities and the
capability machine semantics does not allow capabilities to appear out of thin air.}

\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  If $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}

@@

\subsection{Encapsulation of Local State}
\texttt{\footnotesize{f1}} and \texttt{\footnotesize{f2}} in
\figurename~\ref{fig:prog-f1-and-f2} demonstrate the capability
machine's encapsulation of local state. They are very similar: both
store some local state, call an untrusted piece of code ($\adv$), and
then test whether the local state is unchanged. They differ in the way
they do this. Program \texttt{\footnotesize{f1}} uses our stack-based
calling convention (captured by \texttt{\footnotesize{scall}}) to call
the adversary, so it can use the available stack to store its local
state.  On the other hand, \texttt{\footnotesize{f2}} uses malloc to
allocate memory for its local state and uses an activation-record
based calling convention (described in the technical appendix) to run
the adversarial code.

>>>>

Interestingly, this idea is a variant of a notion of public/private future
worlds that has been previously used in the literature (see
\sectionname~\ref{sec:discussion}). However, temporary regions are new in our setting
and there is an interesting interplay with the recursiveness of the worlds: for
a temporary region, the predicate $H \in \Worlds \rightarrow
\powerset{\MemSegments}$ (which defines the safe memory segments in the
current world) is only required to be monotone w.r.t.\ public
future worlds (i.e.\ safe memory segments remain safe in public future worlds),
but for permanent regions, it must be monotone w.r.t.\ private future worlds.
As a consequence, the memory for a permanent region may not contain local
capabilities (as their safety would be broken in private future worlds), which
in turn implies that only local capabilities may have write-local permission (a
general sanity requirement when using local capabilities\footnote{In fact, local
  capabilities become useless as soon as the adversary has access to a single
  global, write-local capability.})\footnote{For explanation purposes, this
  discussion ignores certain ways to allow for local capabilities in a permanent
  region, for example, by not requiring that they are valid or requiring that
  they are local versions of valid global capabilities.}.
\end{toplas}
\subsection{Worlds}
\label{subsec:worlds}
\begin{figure}[htb]
  \centering
  \includegraphics{w11}

@@

\begin{toplas}
  \section{Malloc and macros}
  \label{sec:malloc-and-macros}
With the semantic model defined, we would like to show its usefulness by proving
the correctness of a series of examples that rely
on well-bracketedness and local-state encapsulation and use the calling
convention to enforce these properties. However, the programs that
run on our capability are assembly programs and even program examples that would
be small in a high-level language, they become big and unintelligible at this
low level. In order to make our program examples intelligible, we introduce a
series of low-level abstractions in the form of macros.

To make the program examples realistic, they have access to a trusted malloc, so
they can allocate new memory. Our results are not specific to
any one implementation of malloc, so rather than providing the implementation,
we provide a specification that malloc must satisfy.

\subsection{Malloc specification}
  \itoplassug{Suggestion: add the malloc specification}
Before we give the malloc specification, we provide some intuition about its requirements.

Malloc is assumed to be part of the trusted computing base, as mentioned in
\sectionname~\ref{sec:capab-mach-with}. A program with access to malloc should
be able to retain this access even if it invokes code it does not trust. In
order to achieve this, a capability for malloc must be global. It is also
important that malloc remains trustworthy to all entities on the machine
throughout execution. That is, malloc should encapsulate its local state and
not depend on any external code. To make sure that malloc cannot be tampered
with directly through the malloc capability, we require it to have enter
permission.

The requirements so far are syntactic and prevent direct tampering in the sense
that the capability intended for invoking malloc only can be used for
invocation. This does, however, not prevent the implementation of malloc from
colluding with other code. Specifically, malloc could depend on the state of a
piece of memory not just under the control of malloc. We also need to make sure
that malloc actually behave as intended. That is when malloc is invoked with a
length argument, then it actually returns a capability for a fresh piece of
memory of that size. It should be fresh in the sense that malloc hasn't given
out a capability for any part of that memory before and won't do it in the
future. Finally, we also need to make sure that malloc behaves correctly when
given a nonsensical length argument such as a negative integer or a capability.
These requirements relate to the behaviour of malloc, so we cannot hope to
express them syntactically. Instead we will use the semantic model defined in
Section~\ref{sec:logical-relation} to express this part of the specification.

In order to express the semantic requirements on malloc, the specification will
require the existence of some initial region for malloc. Malloc does not
disappear from the system, so the malloc region must be permanent. This also
means that the malloc specification should guarantee that use of malloc is safe
at any point of execution no matter who has invoked it. To this end, the
requirement on the initial malloc region should also hold true in any private
future world.
% priv future means that malloc should not depend on local capabilities?

The implementation of malloc should not depend on any memory outside the
implementation of malloc. This means that malloc should not have any
capabilities in its internals that govern other programs. As explained in
\sectionname~\ref{subsec:worlds}, a region's state interpretation function
determines what memory segments it permits with respect to some world. The state
interpretation function of malloc should permit the same regions with respect to
a world with the malloc region as it would with respect to a world that only
contains the malloc region. In other words, the remaining regions in the world
do not influence which regions malloc accepts. Malloc should retain this
property throughout execution, so it must hold true for any private future
region of the initial malloc region.

The malloc specification will also dictate what malloc should do when invoked.
Malloc can be invoked with either a valid or an invalid length argument. In the
case where malloc is invoked with an invalid argument, that is a negative
integer or a capability, we require malloc to simply fail. This part of the
specification does not rely on the malloc region as failing is considered
safe behaviour. When malloc is invoked with a valid length, then malloc should
return a fresh memory segment of the correct length.
\todo[inline]{Everything below this is work in progress..}
Malloc may have been invoked a number of times before any given invocation, so
the internal data structures of malloc may be in a different state than it
started in.
We know that the memory of malloc is in a state consistent with the
protocol dictated by the initial malloc region in the sense that we can assume
that malloc is invoked in a configuration where part of it satisfies a private
future region of the initial malloc region. Knowing the state of the malloc
region allows us to express what we mean by a \emph{fresh} region. This means
that malloc should not keep any capabilities for the freshly allocated region.
To ensure this, we require the footprint of malloc after the call to be disjoint
from the freshly allocated memory. Further, the footprint should satisfy a
public future region of the malloc region the invocation happened in. This means
that it cannot depend on the freshly allocated region. It further allows
subsequent calls to malloc as it provides one of the requirements in the malloc
specification.

Malloc is part of the trusted computing base. If we cannot trust malloc, then we
cannot use the memory it allocates as we have no idea whether it is aliased by
some untrusted program. Because of this assumption, we also require malloc to
promise that the non-argument registers of the register-file remain unchanged
after the call. This allows the caller to keep private capabilities in the
register file. This is strictly speaking not necessary as we can use our calling
convention to protect our local state.

\begin{definition}[Malloc Specification]
  \label{spec:malloc}
  $c_\malloc$ satisfies the specification for malloc iff
  \[  
    \begin{array}{l}
      c_\malloc = ((\entry,\glob),\_,\_,\_) \land\\
      \exists \iota_{\malloc,0} \ldotp \\
      \quad \iota_{\malloc,0}.v = \perma \; \land \\
      \quad \left(
        \begin{array}{l}
          \forall \iota' \futurestr \iota_{\malloc,0}, W,i \ldotp\\
          \quad W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
        \end{array}
        \right) \wedge \\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
        \quad \forall i, n, \size \in \nats \ldotp \forall
        w_{\var{ret}} \in \Words \ldotp \\
        \quad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
        \quad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
        \qquad \memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
        \qquad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
        \qquad \Rightarrow \\
        \qquad \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
        \qquad \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
        \qquad \qquad \quad \Phi \step[j] \Phi' \land \\
        \qquad \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
        \qquad \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
        \qquad \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
        \qquad \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
        \qquad \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
        \qquad \qquad \quad \size - 1 = e'-b'
      \end{array}
      \right) \land\\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp\\
        \quad(\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
        \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
        \qquad \Rightarrow \exists j \in \nats \ldotp \Phi \step[j] \failed
      \end{array}
        \right)
    \end{array}
  \]
\end{definition}
As described previously, the specification of malloc ensure that malloc has no
capabilities pointing \emph{out} of malloc. It does, however, not say anything about
capabilities that point \emph{in} to malloc. We obviously can obviously not
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

Malloc should not just be available to trusted programs, but also to possibly
malicious programs. The specification should therefore entail that the malloc
capability is safe which would make it safe to let even malicious programs have
access to it.
\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  For $c_\malloc$ that satisfies the specification for malloc with region $\iota_{\malloc,0}$,  if $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}

@@

\subsection{Encapsulation of Local State}
\texttt{\footnotesize{f1}} and \texttt{\footnotesize{f2}} in
\figurename~\ref{fig:prog-f1-and-f2} demonstrate the capability
machine's encapsulation of local state. They are very similar: both
store some local state, call an untrusted piece of code ($\adv$), and
then test whether the local state is unchanged. They differ in the way
they do this. Program \texttt{\footnotesize{f1}} uses our stack-based
calling convention (captured by \texttt{\footnotesize{scall}}) to call
the adversary, so it can use the available stack to store its local
state.  On the other hand, \texttt{\footnotesize{f2}} uses malloc to
allocate memory for its local state and uses a calling convention based on heap
allocated activation records (described in the technical appendix) to invoke
the adversarial code.
